Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-01-21 paper.qmd
Added PBMC3k results to discussion

Human behavior often shows a desire for more certainty and a tendency to prefer well-separated views. This emphasizes the importance of clear and distinct clusters. For example, in the UMAP layout of the **PBMC3k** dataset suggested by @chen2023, three distant, well-separated clusters are shown. However, our model reveals that these clusters are actually close to each other in \pD{}. Additionally, the model discovers non-uniform data distribution and non-linear structures within the clusters that are not visible in the UMAP layout, demonstrating the ability of our model in uncovering hidden data characteristics. (added about scDEED in Appendix as well)

Evaluating the error or unexplained variance is important for assessing how well the model fits the data. By examining the error for different numbers of bins, we found that tSNE with a perplexity of $30$ provides a reasonable representation for the **pbmc** dataset. Connecting the closest clusters with line segments in the fitted model further supports the preservation of neighborhood relationships.

>>>>

Our algorithm can also be useful in practice to decide on a reasonable NLDR layout that accurately represents the data, as described with a single-cell gene expression data, **PBMC3k**. The published results of UMAP with default (hyper-)parameter setting by @chen2023 were evaluated. The data showed three clusters, two of which were close. But the published layout showed three well-separated clusters. Therefore, by changing the (hyper-)parameter, we found a better layout with UMAP, which is more representative the data. Additionally, the model discovers non-uniform data distribution and non-linear structures within the clusters that are not visible in the UMAP layout, demonstrating the ability of our model in uncovering hidden data characteristics. Also, as explained in Appendix, with recently introduced approach, scDEED also proved the layout suggested by our layout is better than the published one.
